Supplement: Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2

### 2023-10-01 supplement_figs.Rmd
ready for comment

# GM correlates with Shannon's entropy in HTS datasets

We can think about the scale relationship from an information theoretic point of view. Empirically in most datasets the geometric mean of \(Y^{\parallel}_{n}\) is strongly correlated with Shannon's Entropy \(H\) (supplemental table and supplemental Figure 1) suggesting that  some knowledge of \(W\) is contained in the post-sequencing data that is not strictly compositional.  Intuitively,  underlying systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\).

The logarithm of the geometric mean \( \log(G) \) and \( H \) can be understood from an information theoretic point of view to be an unweighted \( G \) or a weighted measure \( H \) of 'surprisal' in the dataset. Entropy \( H \) and the geometric mean \( G \) (or its logarithm \( \log_2 G \)) are not simple to relate algebraically, but they can be understood in terms of what they are measuring if their description is rephrased in a common language. Recall have a \(D \times N\) matrix of counts \(W\) decomposed into the proportions for the \(n^{th}\) sample \(W^{\parallel}_n\) (or the equivalent probability distribution \( p(w_n) \) ), and its scale \(W^{\perp}_n\).

For notational simplicity assume a single discrete random variable \( X \) with a probability distribution \( p(x) \) over \(1 \dots d \) features. The entropy \( H(X) \) in bits is:

\[
H(X) = -\sum_{i=1}^{d} p_i \log_2 p_i
\]

and for the same distribution log2 of the geometric mean \( G \) is:

\[
\log_2 G = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
\]

As defined here \( H(X) \) is a weighted total of the uncertainty or 'surprisal' contained in \( p(x) \), and relates to the amount of information we would need to have in order to reproduce  \( p(x) \). Conversely, \( G \) is an unweighted mean of the same distribution.  \( G \) is used as the denominator to calculate the centred log-ratio normalization (clr):

\[
clr = \log_2(p_i) - \log_2 G 
\]

over \(i=1 \dots d \).  This form shows explicitly that  each \( p_i \) is compared with the geometric mean \( G \). Thus \( G \) can be interpreted as a measure of difference for how far each \( p_i \) is from \( G \), or in information theoretic terms as an unweighted measure of the  mean surprisal for the distribution.

Thus, we can thus understand \( H(X) \) as a measure of the total weighed surprisal and  \( G \)  as a measure of the average unweighted surprisal for \( p(x) \). The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term.

These two measures are expected to have different behaviours in different distributions of \( p_i \). In the case of a uniform distribution both \( H(X) \) and \( G \) are maximal since \( p(x) \) is equally and identically distributed. Thus, we expect that they are positively correlated here. In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values. In very sparse  datasets, the two measures could become uncoupled because \( H(X) \) could ascribe some uncertainty to the large number of low probability events, while \( G \) would tend to be very small. Here these two measures could be either uncorrelated or exhibit negative correlation. We can see this distributional behaviour in different datasets.

@@

We can see that for most datasets the difference between the \(\overline{G}\) in each group is relatively small. Most significantly, the selex dataset has a very large difference of about 64 fold, and both the 16S and the metatranscriptome dataset have about a 1.5 fold difference. These datasets are candidates for a full scale model correction.

```{r full.scale, echo=F}

@@

plot(data.out[,6],data.out[,1], pch=c(rep(1,19),19), ylim=c(-11, 3),
  xlab='full scale offset', ylab='mean difference between groups')
points(data.out[,6],data.out[,2], pch=c(rep(1,19),19), col='red')
points(data.out[,6],data.out[,3], pch=c(rep(1,19),19), col='orange')
points(data.out[,6],data.out[,4], pch=c(rep(1,19),19), col='cyan')
points(data.out[,6],data.out[,5], pch=c(rep(1,19),19), col='blue')
abline(h=0, col='grey', lty=2)
abline(v=0, col='grey', lty=2)
legend(0.7, 0, legend=c('16S','selex','tscome','metatscome','ss'),
  pch=19, col=c('black','red','orange','cyan','blue'))

@@

The full scale model option allows the investigator to set both the offset between the geometric means of the features and their dispersion. In the offset plot above we can see the cause and the effect of the full model with a fixed gamma of 0.5. We see that the mean location of well centred datasets (yeast transcriptome, single-cell transcriptome and 16S rRNA dataset) are relatively well centred, but could be centred better with small changes in scale ranging from 0 (single cell) to ~0.1 for the 16S and yeast datasets. In contrast, centring the metatranscriptome dataset would require about a 50\% change in relative scale between groups, but as shown in the main text, centring the housekeeping genes is more apt.

The in vitro selection dataset is clearly an outlier in both the difference between the average group geometric mean, and in the offset plot.  However, this dataset can also be used to illustrate the power of the full scale model. In Figure 3 we can see that the default output of ALDEx2 has a centered output. This occurs largely by chance, as the high sparsity of the selected (S) group is balanced almost exactly by the arbitrarily chosen sequencing depth. In this dataset the relative abundances of the majority of features are invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014]. Neither DESeq nor edgeR are able to provide a reasonable analysis of this dataset [@gloorAJS:2016].

A full scale model, where the strong assumption that the mean \(G\) is assumed to be the same between the two groups, dramatically skews the output and the large number of relatively invariant features are now identified as significantly different. While not wrong as long as the assumption that the scales are identical is stated, this is not a useful analysis outcome. Modifying the mean scale difference between groups to be 50-fold different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained without the full scale model. Note that we get exactly the same answer (within random sampling error) with a scale of 0.02 for group 1 and a scale of 1 for group 2, or using a scale of 1 for group 1 and a scale of 50 for group 2. This shows that it is the relative difference between scales that is important, not the absolute values. From this result we can conclude that, on average, the difference in underlying scale in the system is about 50-fold, and this is congruent with the circa 64-fold difference in mean \(G\) between groups. Thus, an advantage of a full scale model is that we have gained information and understanding about the underlying system.

```{r selex, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters. "}

@@

sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,1), conds)
x.1.all <- aldex(selex, conditions=conds, gamma=sel.mu)

sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,120), conds)
x.5.all <- aldex(selex, conditions=conds, gamma=sel.mu)

sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(.00833,1), conds)
x.2.all <- aldex(selex,conditions=conds, gamma=sel.mu)

par(mfrow=c(1,4))
aldex.plot(x.all, main=('gamma=0.5\nmu=G'))
aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,100'))
aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.01, 1'))

@@

Examining the plots, edge R clearly not centred with the median housekeeping functions offset by `r median(meta.edg.qlf$logFC[hk.off==T])` and minimum FDR value  is `r min(p.adjust(meta.edg.qlf$PValue)[hk.off==T])`. Additionally, there is little range in the p-values relative to those seen for DESeq2, and the values are more in line with the range seen with ALDEx2.

>>>>

# GM is related to Information and Shannon's entropy in HTS datasets

Information content is a fundamental property of all systems. We can think about the scale relationship from an information theoretic point of view which measures how much information is encoded in a particular sample. Intuitively,   systems with different scales will contain different amounts of information and so we would expect \(W^{\perp}_n \sim H_n\). Below, we show that the scale of a system as defined by Nixon et al. [-@nixon2023scale] is related to the information content and entropy of the data.

Recall the underlying system is described by a \(D \times N\) matrix of counts \(\mathbf{W}\) decomposed into the proportions for the \(n^{th}\) sample \(\mathbf{W}^{\parallel}_n\) (or the equivalent probability distribution \( \mathbf{p}(w_n) \) ), and its scale \(\mathbf{W}^{\perp}_n\). Sequencing  measures which is related to the underlying proportion; i.e., \(\mathbf{Y}^{\parallel}_n \sim \mathbf{W}^{\parallel}_n\)

For notational simplicity assume we will use a single discrete random variable to represent the probability distribution; i.e. \( X =  \mathbf{p}(w_n) \)  over \(1 \dots d \) features. The geometric mean can be calculated as:

\[
\log_2 G(X) = \frac{1}{d} \sum_{i=1}^{d} \log_2 p_i
\];

here every element in \(log_2 G(X)\) is the logarithm of the elemental probability \( log_2(p_i)\).

In information theory, the elemental amount of information or uncertainty for \(p_i\) is the inverse \(I(X)_i = -log_2(p_i)\)[@rezaInfoIntro]. The total entropy of the system  \( H(x) \) is the weighted sum of the elemental information

\[
H(x) = -\sum_{i=1}^{d} p_i \log_2 p_i
\]

In this formulation, \( \log_2(G(X)) \) and \( H(X) \) can be understood from an information theoretic point of view to be an unweighted and a weighted measure  of 'surprisal' in the dataset.  The total and the mean surprisal are related by the number of terms in \( p(x) \) and by the weighting factor for each term. \( H(X) \) corresponds to the amount of information needed in order to reproduce  \( X \). Conversely, \( G \) is an unweighted mean of the same distribution and  \( G \) is the denominator to calculate the centred log-ratio normalization (CLR).

Empirically, we can see that \(G(\mathbf{Y}^{\parallel}_{n})\) is strongly correlated with Shannon's Entropy \(H(Y^{\parallel}_{n})\) (Supplemental Figure 1 and Table 1).  These two measures are expected to have different behaviours with different distributions of \( p_i \). In the case of a uniform distribution both \( H(Y^{\parallel}_{n}) \) and \( G(Y^{\parallel}_{n}) \) are maximal since \( p(x) \) is equally and identically distributed. Thus, we expect that they are positively correlated here. In a Normal or a skewed distribution, we also anticipate a positive correlation because both are affected in the same direction by outlier values. In very sparse  datasets, the two measures could become uncoupled because \( H(Y^{\parallel}_{n}) \) could ascribe some uncertainty to the large number of low probability events, while \( G(Y^{\parallel}_{n}) \) would tend to be very small. Here these two measures could be either uncorrelated or exhibit negative correlation. We can see this distributional behaviour in different datasets.

@@

We can see that for most datasets the difference between the \(\overline{G}\) in each group is relatively small. Most significantly, the selex dataset has a very large difference of about 100-fold, and both the 16S and the metatranscriptome dataset have about a 1.5 fold difference. These three datasets are candidates for a full scale model correction.

```{r full.scale, echo=F, fig.cap="Plot of the offset of the mean differnece between groups as a function of scale ratio. For this, the default scale of 1:1 was altered in increments of 0.1 keeping the gamma parameter (dispersion) at 0.5. The filled circle shows the outcome when the calculation is done using the geometric mean and the same gamma parameter."}

@@

plot(data.out[,6],data.out[,1], pch=c(rep(1,19),19), ylim=c(-11, 4),
  xlab='full scale offset', ylab='mean difference between groups')
points(data.out[,6],data.out[,2], pch=c(rep(1,19),19), col='red')
points(data.out[,6],data.out[,3], pch=c(rep(1,19),19), col='orange')
points(data.out[,6],data.out[,4], pch=c(rep(1,19),19), col='cyan')
points(data.out[,6],data.out[,5], pch=c(rep(1,19),19), col='blue')
abline(h=0, col='grey', lty=2)
abline(v=0, col='grey', lty=2)
legend(0.7, 0, legend=c('16S','selex','tscome','metatscome','ss'),
  pch=19, col=c('black','red','orange','cyan','blue'))

@@

Nixon et al. [-nixon2023scale] showed that many operations on HTS datasets relied on both the proportion and scale components of the data. Moreover, all normalizations impose a scale model on the data, but the appropriateness of these models has never been explicitly acknowledged or tested. Thus, the full scale model option allows the investigator to set both the offset between the geometric means of the features and their dispersion and observe how this affects the analysis outcome.

In the offset plot above we can see the cause and the effect of the full model with a fixed gamma of 0.5 and a base scale of 1 in each group. We see that the mean location of well centred datasets (yeast transcriptome, single-cell transcriptome) are close to 0, but could be centred better with small changes in scale ranging from 0 (single cell) to 1:1.1 for the yeast transcriptome dataset. In contrast, centring the 16S dataset requires about a 1:1.3 fold change. The metatranscriptome dataset would require about a 0.5:1 change in relative scale between groups, but as shown in the main text, centring the housekeeping genes is more apt.

The in vitro selection dataset is clearly an outlier in both the difference between the average group geometric mean, and in the offset plot.  However, this dataset can  be used to illustrate the power of the full scale model. In Figure 3 we can see that the default output of ALDEx2 has a centered output. This occurs largely by chance, as the high sparsity of the selected (S) group is balanced almost exactly by the arbitrarily chosen sequencing depth so the non-selected group (NS) appears to have a similar location as the S group. The difference in entropy between the two groups and the differences in geometric mean are very large, with the difference in \(\log_2{\overline{G}(S)}\) and \(log_2{\overline{G}(NS)}\) being about \(2^{6.6}\). Setting the scale of both the S and NS groups to 1 we find that the difference in location is approximately \(2^{7.7}\) in close agreement with the difference the geometric means. For this dataset to be centered we need to have a scale ratio \( \approx \) 1:50 or more. In fact, in this dataset the relative abundances of the majority of features are nearly invariant, but this is masked by the large absolute changes in a small number of features [@mcmurrough:2014], thus changing the scale of the data. Neither DESeq nor edgeR are able to provide a reasonable analysis of this dataset because the normalizations used assume equivalent scales [@gloorAJS:2016].

Figure 3 shows an effect plot of various scale models with this dataset. The full scale model, where the strong assumption that the mean \(G\) is assumed to be 1:1 between the two groups, dramatically skews the output and the large number of relatively invariant features are now identified as significantly different. While not wrong as long as the assumption that the scales are identical is stated, this is not a useful analysis outcome. Modifying the mean scale difference between groups to be \( \approx \) 1:50 different moves the centre of the large number of relatively invariant features to the centreline of no difference, and recapitulates the default result obtained using \(G_n\) as the scale estimate. Note that we get exactly the same answer (within random sampling error) with a scale of 0.02 for group 1 and a scale of 1 for group 2, or using a scale of 1 for group 1 and a scale of 50 for group 2. This shows that it is the relative difference between scales that is important in this dataset, not the absolute values. From this result we can conclude that, on average, the difference in underlying scale in the system is about \(\approx 50\)-fold, and this is congruent with the circa 100-fold difference in  \(\overline{G}\) between groups; the discrepancy being explained because the default scale model is applied uniformly to all samples, whereas the different values of the within-group geometric means ranging over a \{ > 2.5 \) fold range.  Thus, an advantage of a full scale model is that we have gained both information and understanding about the drivers of asymmetry underlying system.

```{r selex, echo=F, warning=F, message=F,comment=F, fig.cap="Effect plots of the selex dataset with various gamma and scale parameters. All scales are calculated with a logNormal distribution to ensure symmetry for the user. "}

@@

sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(10,10), conds)
x.1.all <- aldex(selex, conditions=conds, gamma=sel.mu)

# setting ratio as 1:50 explicitly
sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(1,50), conds, log=FALSE)
x.5.all <- aldex(selex, conditions=conds, gamma=sel.mu)

# setting ratio in log space
# 2^-5.64 : 2^0  = 0.02 : 1
sel.mu <- aldex.makeScaleMatrix(gamma=0.5, mu=c(-5.64,0), conds, log=TRUE)
x.2.all <- aldex(selex,conditions=conds, gamma=sel.mu)

par(mfrow=c(1,4))
aldex.plot(x.all, main=('gamma=0.5\nmu=G'))
aldex.plot(x.1.all, main=('gamma=0.5\nmu=1,1'))
aldex.plot(x.5.all, main=('gamma=0.5\nmu=1,50'))
aldex.plot(x.2.all, main=('gamma=0.5\nmu=0.02, 1'))
```

@@

Examining the plots, edge R clearly not centred with the median housekeeping functions offset by `r median(meta.edg.qlf$logFC[hk.off==T])` and minimum FDR value  is `r min(p.adjust(meta.edg.qlf$PValue)[hk.off==T])`. Additionally, as shown in Figure 4there is little range in the p-values relative to those seen for DESeq2, and the values are more in line with the range seen with ALDEx2.
